\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-11-06 casestudies.tex
a la carte

The elimination of inductive types defined via \lsti{FInductive} is
mostly via the \lsti{FRecursion} and \lsti{FInduction} commands.
An exception is seemingly trivial ``inversion lemmas''.
For example, consider the lemma \lsti!\forall t, \neg step tm_true t!
stating that \lsti!tm_true! is irreducible.
If \lsti{step} were an ordinary inductive type, then it could be proved in Coq
with \lsti{intros t H; inversion H}.
But since \lsti{step} is extensible, it seems that the lemma has to be
proved by \lsti{FInduction} on \lsti{step}, and that the programmer
has to manually verify that the lemma still holds in derived families
that add new constructors to \lsti{step}.
%
However, we observe that it is lighter-weight to use overriding: 
the programmer can specify that the proof of the lemma should be overridden
in any derived family that extends \lsti{step} with new constructors,
and in return, they are permitted to treat \lsti{step} as an ordinary
inductive type in the proof and thus discharge it with \lsti{inversion},
with the plugin automatically trying the same proof script in a derived
family to override the proof.

>>>>

Elimination of inductive types defined via \lsti{FInductive} is
mostly via the \lsti{FRecursion} and \lsti{FInduction} commands.
An exception is some seemingly trivial ``inversion lemmas''.
For example, consider the lemma \lsti!\forall t, \neg step tm_true t!
stating that \lsti!tm_true! is irreducible.
If \lsti{step} were an ordinary inductive type, then it could be proved in Coq
with \lsti{intros t H; inversion H}.
But since \lsti{step} is extensible, it seems that the lemma has to be
proved by \lsti{FInduction} on \lsti{step}, and that the programmer
has to manually verify that the lemma still holds in derived families
that add new constructors to \lsti{step}.
%
However, we observe that it is lighter-weight to use overriding (\cref{sec:override}):
the programmer can specify that the proof of the lemma should be overridden
in any derived family that further binds \lsti{step},
and in return, they are permitted to treat \lsti{step} as an ordinary
inductive type in the proof and thus discharge the proof with \lsti{inversion}.
The plugin then automatically tries the same proof script in a derived
family to override the proof.

We note that the current plugin implementation does not support
mutually inductive types or mutual induction, which will be
useful for modeling languages much more complex than STLCs.
We believe that generalizing our design to mutually inductive types does
not pose theoretical challenges, however.

### 2022-11-06 coqimpl.tex
a la carte

In family \lsti{STLC},
field \lsti{env} and the case handlers for \lsti{subst}
are not further-bindable by derived families.
In contrast, \lsti{tm}, \lsti{subst}, and the related equalities
can be further bound.
The distinction is reflected in the translations.
Further-bindable fields are translated to module types that export only types of the fields.
Non-further-bindable fields are translated to modules that export definitional equalities
on the fields.

>>>>

In family \lsti{STLC},
field \lsti{env} and the case handlers for \lsti{subst}
are not further-bindable by derived families.
In contrast, \lsti{tm}, \lsti{subst}, and the related equalities
can be further bound.
The distinction is reflected in the translations.
The further-bindable fields are translated to module types that export only types of the fields.
The non-further-bindable fields are translated to modules that export definitional equalities
on the fields.
Opaque definitions can be further bound, too (\cref{sec:override});
they are translated to modules that export opaque definitions.

### 2022-11-06 lang-design.tex
a la carte

@@

@@

\noindentparagraph{Overriding.}

In an OO language, a subclass can override methods of a superclass.
Similar expressiveness is useful for metatheory mechanization, too.
%
For example, in a derived family, the programmer may want to override
the proof of a lemma, using a different proof strategy than it was proved
in the base family.
%
Overriding is potentially incompatible with having equalities on late bound names, however.

Coq distinguishes \emph{opaque} definitions from \emph{transparent} ones.
\Lang supports the overriding of opaque definitions, which
include most proofs.
It is safe to override opaque definitions, because Coq will never attempt to unfold them.
For transparent definitions, the common case is that the programmer does
not want to override them.
In \cref{fig:stlc-mechanized}, "env", "empty", "subst", and "steps" are
transparent---that is, they are not defined with \lsti{Qed}.
\Lang treats transparent definitions as non-overridable by default.
Thus, the definitional equalities on "env", "empty", and "steps",
as well as the propositional equality on "subst", are available to the
type checker for type-checking the families.

\Lang does allow the overriding of transparent definitions that are
explicitly marked as \lsti{Overridable} by the programmer.
Overriding is made safe by requiring that when an overridable field is
overridden, code whose type checking involves unfolding that field
should be overridden too.
We expect this feature to be used only occasionally.
Our case studies contain uses of overriding (\cref{sec:casestudies}).

\noindentparagraph{Sound logical reasoning.}

In \cref{fig:stlc-mechanized}, just because the "typesafe" theorem in
"STLC" is inherited and reused by "STLCFix", it does not follow that
in "STLCFix" the programmer can use "typesafe" to prove "progress".
If the programmer did, then they would be committing the logical fallacy
of circular reasoning: the proof of "progress" would depend on "typesafe",
yet the proof of "typesafe" depends on "progress".
%
Such circularity would easily lead to logical inconsistency:
\YZ{Mention non-exhaustivity/overriding already poses challenges to ensuring consistency.}

@@

Note that the requirement still allows a derived family to introduce
new declarations into the context of a field. \EDJ{Maybe a better wording is ``the relative order between fields are preserved during inheritance''?}
For example, the left column of \cref{fig:stlc-isorec-prod} shows an
extension of STLC with iso-recursive types, where "tysubst" is
introduced into the context of "hasty".

@@

\noindentparagraph{Composing families.}

Families can be readily reused to construct larger extensions that
\emph{mix in}~\cite{mixin-1990} the functionalities of the individual
families.
In \cref{fig:stlc-isorec-prod}, family "STLCFixIsorec" is declared as
an "STLC" extension that mixes in "STLCFix" and "STLCIsorec".
As a result, \lsti{STLCFixIsorec.typesafe} is readily the proof of the type-safety
theorem of an STLC equipped with fixpoints and iso-recursive types.\EDJ{I feel this paragraph will cause more confusion---because it is inheritance judgement/linkage transformer being mixed in, not family itself. Conceptually not distinguishing family and linkage transfomer will make reader wonder what this magic black box here actually is}

@@

>>>>

\subsection{Extensible Inductive Types and Exhaustive Recursion/Induction}

@@

\subsection{Late Binding and Equalities}
\label{sec:latebind}

@@

\subsection{Overriding}
\label{sec:override}
%\noindentparagraph{Overriding.}

In an OO language, a subclass can override methods of a superclass.
Similar expressiveness is useful for metatheory mechanization, too.
%
For example, in a derived family, rather than adding new cases to an
induction proof, the programmer may prefer overriding the proof
entirely, as we observe in our case studies (\cref{sec:casestudies}).
%
Overriding is potentially incompatible with having equalities on late bound names, however.

Coq distinguishes \emph{opaque} definitions from \emph{transparent} ones.
\Lang supports the overriding of opaque definitions, which
include most proofs.
It is safe to override opaque definitions, because the type checker will
never attempt to unfold them.
For transparent definitions, the common case is that the programmer does
not want to override them.
In \cref{fig:stlc-mechanized}, "env", "empty", "subst", and "steps" are
transparent---that is, they are not defined with \lsti{Qed}.
\Lang treats transparent definitions as non-overridable by default.
Thus, the definitional equalities on "env", "empty", and "steps",
as well as the propositional equality on "subst", are available to the
type checker for type-checking the families.

\Lang does allow the overriding of transparent definitions
explicitly marked as \lsti{Overridable} by the programmer.
Overriding is made safe by requiring that when an overridable field is
overridden, code whose type checking involves unfolding that field
should be overridden too.
We expect this feature to be used occasionally.

\subsection{Sound Logical Reasoning}
\label{sec:sound}

%\noindentparagraph{Sound logical reasoning.}

In \cref{fig:stlc-mechanized}, just because the "typesafe" theorem in
"STLC" is inherited and reused by "STLCFix", it does not follow that
in "STLCFix" the programmer can use "typesafe" to prove "progress".
If the programmer did, then they would be committing the logical fallacy
of circular reasoning: the proof of "progress" would depend on "typesafe",
yet the proof of "typesafe" depends on "progress".
%
Such circularity would easily lead to logical inconsistency:
%\YZ{Mention non-exhaustivity/overriding already poses challenges to ensuring consistency.}

@@

Note that the requirement still allows a derived family to introduce
new declarations into the context of a field. %\EDJ{Maybe a better wording is ``the relative order between fields are preserved during inheritance''?}
For example, the left column of \cref{fig:stlc-isorec-prod} shows an
extension of STLC with iso-recursive types, where "tysubst" is
introduced into the context of "hasty".
%
In the event that it cannot be inferred where to place of a newly introduced field,
programmer annotation is required.

@@

\subsection{Composing Families as Mixins}

%\noindentparagraph{Composing families.}

Families can be readily reused to construct larger extensions that
\emph{mix in}~\cite{mixin-1990} the functionalities of the individual
families.
A family like \lsti{STLCFix} can be viewed as a family-to-family functor---%
and hence a mixin, in the sense of \cite{flatt1998mixin}---that
transforms any family providing the base STLC functionalities into a new
family additionally supporting fixpoints.
%\EDJ{I feel this paragraph will cause more confusion---because it is inheritance judgement/linkage transformer being mixed in, not family itself. Conceptually not distinguishing family and linkage transfomer will make reader wonder what this magic black box here actually is}

In \cref{fig:stlc-isorec-prod}, family "STLCFixIsorec" is declared as
an "STLC" extension that mixes in "STLCFix" and "STLCIsorec".
The family is declared with minimal verbiage, yet
\lsti{STLCFixIsorec.typesafe} is automatically a proof of the type-safety
theorem of an STLC equipped with fixpoints and iso-recursive types.

@@

%\YZ{
%Family IDENT.
%
%Family IDENT extends IDENT_0.
%
%Family IDENT extends IDENT_0 using IDENT_1, IDENT_2, ..., IDENT_n.
%
%Semantically, family IDENT is the composition λ IDENT_0. IDENT_n(…(IDENT_2(IDENT_1(IDENT_0)))…).
%
%IDENT_0 is called the immediate superfamily of IDENT.
%
%If the using-clause is present, then for each i > 0, IDENT_0 must be the same as, or a derived family of, the immediate superfamily of IDENT_i.
%
%If there is any name conflict among nested fields, then either the programmer has to resolve it or the composition is rejected upon attempting to close the family.
%}

\subsection{Injectivity and Disjointness of Constructors via Partial Recursors}

### 2022-11-06 related-work.tex
a la carte

This section discusses related work.

@@

>>>>

There exist many approaches to modular mechanization or proof reuse,
with different focuses and trade-offs.

\noindentparagraph{\`A la carte.}

The design pattern \emph{data types à la carte} (DTC) introduced for Haskell by~\citet{dtc}
is a source of inspiration for many approaches
\cite{delaware2013,delaware2013b,schwaab2013modular,keuchel2013generic,forsta2020}.
The original DTC encoding requires type-level general recursion that
fails the strict positivity check imposed by proof assistants including Coq and Agda.
\citet{delaware2013} introduce \emph{meta-theory à la carte} (MTC):
it overcomes the problem by using Church encodings for
data types and using Mendler-style folds for evaluation,
though it requires \lsti{Set} impredicativity.
Feature composition is automated through heavy use of type classes.
The framework is implemented as a Coq library.
\citet{schwaab2013modular} adapt DTC to Agda by considering a restricted
class of functors that admit least fixed points.
\citet{keuchel2013generic} use datatype-generic programming techniques
for the underlying representation of type-level fixed points and
avoid \lsti{Set} impredicativity.

All of these approaches are largely extra\-linguistic, in that they work
within the confine of the language offered by a proof assistant, which
comes with trade-offs.
On the one hand,
they can be conveniently distributed as libraries, and the encoding can
be more easily adapted for new purposes.
For example, MTC~\cite{delaware2013} has been applied to implementing
composable program adverbs~\cite{liwei2022} and been adapted to support
unanticipated feature extensions that require type changes
\cite{delaware2013b}.

On the other hand, the extra\-linguistic nature of the approaches tend
to lead to non-idiomatic code and offset their user-friendliness.
In particular, because data types have to be encoded (rather than expressed
through natively supported inductive types), the resulting code is
less readable and the resulting programming style inaccessible to non-experts.
In addition, extra programmer effort may be required, such as
having to manually prove additional well-formedness conditions on the algebras.

\citet{forsta2020} introduce \emph{Coq à la carte}.
It still follows DTC, but rather than embracing DTC's use of generic
fixed points, it considers specific instantiations instead.
The resulting mechanism appears more streamlined than prior \emph{à la
carte} approaches particularly for its extensive tool support for generating boilerplate code.
%
The approach does not seem to fully address the expression problem,
as \citet{liwei2022} observe, in that extending a composed
mechanization requires recompilation.

Our approach addresses the expression problem by extending the
linguistic facilities offered by a proof assistant.
Families, in particular, offer an organizational advantage.  They allow
grouping and coevolving related types, functions, and proofs without
explicit parameterization;
all further-bindable fields and all opaque fields are automatically
extensibility hooks.
Because family polymorphism does not require explicit parameterization
or complex encodings,
the resulting programming experience and code are accessible to the
working Coq programmer.
The more OO aspects of family polymorphism, such as the ability to
override and families as mixins, also facilitate extensibility and reuse
with minimal programmer effort.

\bigskip

@@

\ifShowOldWriting

### 2022-11-06 stlc-isorec-prod.tex
a la carte

\begin{lstlisting}
Family STLCFixIsorec extends STLC[@\tanc{STLCIsorecfix-start}@]
using STLCFix, STLCIsorec.[@\hfill\codecomment{STLC extended}@]
[@\hfill\codecommentend{with fixpoints and iso-recursive types}@]
End STLCFixIsorec.[@\tanc{STLCIsorecfix-end}@]
\end{lstlisting}

>>>>

\begin{lstlisting}
Family STLCFixIsorec extends STLC[@\tanc{STLCIsorecfix-start}@]
using STLCFix, STLCIsorec.[@\hfill\codecommentstart{STLC extended}@]
[@\hfill\codecommentend{with fixpoints and iso-recursive types}@]
End STLCFixIsorec.[@\tanc{STLCIsorecfix-end}@]
\end{lstlisting}
